\documentclass[12pt]{article}
\usepackage{amsmath,amssymb}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage[hidelinks]{hyperref}
\setlength{\parindent}{0pt}
\usepackage{xcolor}
\usepackage{etoolbox}  %Supports toggle commands
\definecolor{darkblue}{RGB}{0,0,139}

%SETTINGS
\newtoggle{solutions} \toggletrue{solutions}

\begin{document}

\begin{center}
\textbf{Midterm Exam 1} \\
Introduction to Health Economics: EC 387 \\
Boston University \\
Spring 2026 \\
Professor Pauline Mourot
\end{center}

\vspace{1em}

[ADD DISCLAIMER EXAM]

\noindent \textbf{PART I: TRUE/FALSE/UNCERTAIN [56 points]}\\

For each of the questions below, please explain your answer.
The quality of the explanation is an important part of the grade for these problems and
can also help you earn substantial partial credit.
In answering each question, 1-3 sentences of explanation should be sufficient.

\vspace{7mm}

[ADD 5 TFU QUESTIONS, 7 questions total at 8 points each]


\noindent \textbf{QUESTION}\\
A risk-averse individual has a lottery ticket that would give them \$0 with 50\% probability and \$1,000 with 50\% probability.
They would prefer to take \$500 (for sure) instead of keeping the lottery ticket. 

\iftoggle{solutions}{
{\color{darkblue}

\vspace{5mm}

TRUE

}}{}

\vspace{7mm}

\noindent \textbf{QUESTION}\\
A risk-averse individual has a lottery ticket that would give them \$0 with 50\% probability and \$5,000 with 50\% probability.
They would prefer to take \$2,400 (for sure) instead of keeping the lottery ticket.

\iftoggle{solutions}{
{\color{darkblue}

UNCERTAIN - Not enough information is provided.

}}{}

\vspace{7mm}


\vspace{10mm}

\noindent \textbf{PART II: Quantitative Exercises [44 points]}\\

\vspace{5mm}

\noindent \textbf{Exercise 1 [12 points]}\\

Tony faces uncertainty about his health state next year.
There is a 80\% chance he will be perfectly healthy and have \$0 in medical bills,
but there is a 20\% chance he will get sick and have \$4,000 in medical bills.
His current wealth is \$12,000. His preferences are represented by the utility function

\begin{equation*}
U(w) = \sqrt{w}.
\end{equation*}

\medskip

\noindent \textbf{(a)} If he does not have insurance, what is Tony's expected wealth next year?

\iftoggle{solutions}{
{\color{darkblue}
\begin{align*}
E[w] &= 0.8 \times 12000 + 0.2 \times (12000 - 4000)\\
& = 11,200
\end{align*}
}}{}

\medskip

\noindent \textbf{(b)} If he does not have insurance, what is Tony's expected utility next year?
(round to the nearest decimal place)

\iftoggle{solutions}{
{\color{darkblue}
\begin{align*}
E[U(w)] & = 0.8 \times \sqrt{12000} + 0.2 \times \sqrt{12000 - 4000}\\
& = 105.5
\end{align*}
}}{}

\medskip

\noindent \textbf{(c)} If he does not have insurance,
what is Tony's certainty equivalent next year?
(round to the nearest decimal place)

\iftoggle{solutions}{
{\color{darkblue}
\begin{align*}
U(CE) &= E[U(w)] \\
\sqrt{CE} &= 105.5 \\
CE &= 11,130.3
\end{align*}
}}{}

\medskip

\noindent \textbf{(d)} What would Tony's certainty equivalent be under a full insurance plan with a premium of \$3,000?

\iftoggle{solutions}{
{\color{darkblue}
\begin{equation*}
CE = 12000 - 3000 = 9000
\end{equation*}
}}{}

\medskip

\textbf{(e)} Would Tony be willing to purchase a full insurance plan with a premium of \$2,000?

\iftoggle{solutions}{
{\color{darkblue}
Expected utility if purchase is

\begin{equation*}
U = \sqrt{10000} = 100 < 105.5,
\end{equation*}

so no.
}}{}

\medskip

\textbf{(f)} What would the actuarially fair premium for a full insurance plan for Tony be?

\iftoggle{solutions}{
{\color{darkblue}
\begin{equation*}
0.8 \times 0 + 0.2 \times 4000 = 800
\end{equation*}
}}{}


\vspace{5mm}

\noindent \textbf{Exercise 2 [24 points]}\\
Assume that an individual has income of \$80,000. The individual faces a risk of hospitalization,
which occurs with probability $p = 0.1$.
If this happens, the total cost will be \$15,000.
The individual's utility over income net of out-of-pocket medical costs is

\begin{equation*}
u(y) = \sqrt{y}.
\end{equation*}

Before facing the risk, the individual is deciding between two health insurance plans (Plan A and Plan B)
and remaining uninsured.

\medskip

\textbf{Plan A:}  
The individual pays a premium of \$1,050 (whether she shows up in the hospital or not) for a health insurance plan
with a deductible of \$5,000.
After paying out-of-pocket costs up to that deductible, the individual is fully insured.

\medskip

\textbf{Plan B:}  
The individual pays a premium of \$1,560 (whether she shows up in the hospital or not) for a health insurance plan
with no deductible and full insurance (so there are no out-of-pocket costs at all).

\medskip

\textbf{(a)} From the perspective of the firm, compute the expected cost and the expected profit to the insurer
from selling each plan.

\iftoggle{solutions}{
{\color{darkblue}
Firm expected costs are how much they should expect to pay out. In the fully insured plan, this aligns with the consumer’s expected cost. In the plan with a deductible, the firm pays less than the full amount because of the deductible. Expected profits are the premiums minus the expected costs.

Plan A:

\begin{align*}
\text{Expected costs} &= 0.1 \times (15000 - 5000) = 1000 \\
\text{Expected profits} &= 1050 - 1000 = 50
\end{align*}

Plan B:

\begin{align*}
\text{Expected costs} &= 0.1 \times 15000 = 1500 \\
\text{Expected profits} &= 1560 - 1500 = 60
\end{align*}
}}{}

\medskip

\textbf{(b)} Does the consumer prefer to purchase Plan A, to purchase Plan B, or to remain uninsured?

\iftoggle{solutions}{
{\color{darkblue}
To calculate consumer preferences, we use the utility function to calculate the consumer's expected utility
under each of the plans.

Uninsured:
\begin{equation*}
E[U] = 0.1 \times \sqrt{80000 - 15000} + 0.9 \times \sqrt{80000} = 280.0535
\end{equation*}

Plan A:
\begin{equation*}
E[U] = 0.1 \times \sqrt{80000 - 5000 - 1050} + 0.9 \times \sqrt{80000 - 1050} = 280.0761
\end{equation*}

Plan B:
\begin{equation*}
E[U] = \sqrt{80000 - 1560} = 280.0714
\end{equation*}

The consumer prefers to insure themselves. Further, they prefer Plan A to Plan B.
}}{}

\medskip

\textbf{(c)} Does the answer to part (b) change if the consumer is risk-neutral?

\iftoggle{solutions}{
{\color{darkblue}
Now consider a risk-neutral utility function:

\begin{equation*}
u(y) = y.
\end{equation*}

Uninsured:
\begin{equation*}
E[U] = 0.1 \times (80000 - 15000) + 0.9 \times 80000 = 78500
\end{equation*}

Plan A:
\begin{equation*}
E[U] = 0.1 \times (80000 - 5000 - 1050) + 0.9 \times (80000 - 1050) = 78450
\end{equation*}

Plan B:
\begin{equation*}
E[U] = 80000 - 1560 = 78440
\end{equation*}

Now the consumer prefers being uninsured to paying for an insurance plan. Intuitive explanations here are also accepted.
}}{}

\medskip

\textbf{(d)} Suppose the government decides to make all high-deductible plans (such as Plan A) illegal.
If Plan A is eliminated, would the consumer prefer Plan B or prefer remaining uninsured?
Provide intuition for your answer in 1–3 sentences.

\iftoggle{solutions}{
{\color{darkblue}
Using the same calculations from part (b), we can now see that consumers will prefer Plan B to being uninsured.
The consumer is still risk averse, so they are willing to pay for certainty rather than remaining uninsured.
Their level of risk aversion means that they are willing to still bear some risk and purchase a plan with a deductible,
but if that option is not available, they still prefer the less desirable insurance plan to being uninsured.
Their total willingness to pay is higher than the premium on Plan B.
}}{}


\vspace{10mm}



\noindent \textbf{Exercise 3 [8 points]}\\
This problem works through a simplified version of the Grossman model.
Individuals have the following utility function for consumption, $U(c)=c$.
Consumption is purchased from labor earnings, $y$, which is product of the individual's hourly wage 30 and total hours worked,
$T^w$. In other words,
\begin{equation*}
c = y = 30 \times T^w .
\end{equation*}

Total hours available is 24, and must be split between work hours $T^w$, hours spent sick $T^s$,
and hours spent producing health $T^H$.
Health, $H$, is produced solely from time spent producing health (e.g., exercise); i.e., $H = T^H$
and $T^S = 24 \times \exp(-H) = 24 \times \exp(-T^H)$.

\medskip

\noindent \textbf{(a)} What's the maximum utility the individual can achieve?\\
\textit{(Hint: $\log(24) \approx 3.18$.)}

\iftoggle{solutions}{
{\color{darkblue}

\vspace{5mm}

\begin{align*}
c &= y = 30 \times T^w\\
& = 30 \times (24 - T^S - T^H) \\
  &= 30 \times (24 - 24 \exp(-T^H) - T^H)
\end{align*}

Taking the first-order condition with respect to $T^H$:

\begin{align*}
30 \times (24\exp(-T^H) - 1) &= 0 \\
\exp(-T^H) &= \frac{1}{24}\\
\frac{1}{\exp(T^H)} &= \frac{1}{24} \\
\exp(T^H) &= 24 \\
T^H &= \log(24)\\
T^H &\approx 3.18
\end{align*}

and 

\begin{align*}
    T^S &= 24 \exp(-T^H)\\
    &= 24 \frac{1}{\exp(\log(24))}\\
    & = 24 \times \frac{1}{24} \\
    &= 1
\end{align*}

and

\begin{align*}
T^w &= 24 - T^S - T^H \\
&= 19.82 \\
\end{align*}

and finally 
\begin{align*}
U(c) &= c = 30 \times T^w \\
&= 594.6
\end{align*}

}}{}

\medskip

\noindent \textbf{(b)} What's the allocation of time when the individual achieves the maximum utility level?

\iftoggle{solutions}{
{\color{darkblue}

\vspace{5mm}

See derivations above:

\begin{align*}
T^S &= 1 \\
T^H &= \log(24) = 3.18 \\
T^w &= 24 - T^S - T^H = 19.82
\end{align*}

}}{}


\end{document}