\documentclass[12pt]{article}
\usepackage{amsmath,amssymb}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage[hidelinks]{hyperref}
\setlength{\parindent}{0pt}
\usepackage{xcolor}
\usepackage{etoolbox}  %Supports toggle commands
\definecolor{darkblue}{RGB}{0,0,139}
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

%SETTINGS
\newtoggle{solutions} \togglefalse{solutions}

\begin{document}

\begin{center}
\textbf{Problem Set 4} \\
Introduction to Health Economics: EC 387 \\
Boston University \\
Spring 2026 \\
Professor Pauline Mourot
\end{center}

\vspace{1em}

\noindent \textbf{Exercise 1: TRUE/FALSE/UNCERTAIN}\\

For each of the questions below, please explain your answer.
The quality of the explanation is an important part of the grade for these problems and
can also help you earn substantial partial credit.
In answering each question, 1-3 sentences of explanation should be sufficient.

\vspace{7mm}

\noindent \textbf{QUESTION 1}\\
In the Rotschild and Stiglitz (1976) model of insurance markets with asymmetric information,
there always exist a separating equilibrium in which the low-risk consumers buy a different insurance contract than the high-risk consumers.

\iftoggle{solutions}{
{\color{darkblue}

\vspace{5mm}

FALSE.
In the Rotschild and Stiglitz (1976) model, there may not exist a separating equilibrium.
A separating equilibrium is less likely
when the low-risk population is large compared to the whole population. 
}}{}

\vspace{7mm}

\noindent \textbf{QUESTION 2}\\
In the Rotschild and Stiglitz (1976) model of insurance markets with asymmetric information,
the existence of high-risk consumers make the low-risk consumers worse off compared to the full information case.

\iftoggle{solutions}{
{\color{darkblue}

\vspace{5mm}

TRUE.
When the insurer cannot tell the consumer types apart,
the contract that the low-risk consumers would buy in the perfect information case would attract high-risk consumers, so the insurer would not offer that contract.
When a separating equilbrium exists,
the insurer would offer a contract that is more expensive and less generous than the one the low-risk consumers would buy in the perfect information case.
This contract lies on a lower indifference curve compared to the full information contract,
so the low-risk consumers are worse off than in the full information case.

}}{}

\vspace{7mm}

\noindent \textbf{QUESTION 3}\\
The fact that health insurers try to attract healthier consumers with contract features is called ``cream skimming'',
and risk-adjustment is one way to try to prevent it.

\iftoggle{solutions}{
{\color{darkblue}

\vspace{5mm}

TRUE.

}}{}

\vspace{7mm}


\noindent \textbf{Exercise 2: A selection market}\\

\medskip



\end{document}